What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-06-26 paper/main.tex
def 2.4 fn to relation
typos/minor changes

\subsection{Timeline}\label{sec:timeline}
We present timelines as graphic visualizations. It contains the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start''.
    \item Every node represents one time step, and each node has a propositional logic formula $\Psi$, which specifies the behavior of atomic propositions at that time step. The formula $\Psi$ must be true at that time step. If $\Psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item A node in egg shape with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting point.
    \item Timelines can be in parallel, signifying that either of the parallel timeline could happen.
\end{itemize}

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualizations works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure*}

@@

\paragraph*{Rule-based simplification}
Here we show a demonstrating subset of simplification rules we encoded. In theory one could add more rules to the tool, so long as they are sound; here we only choose to encode the rules that represent common patterns we have observed in the generated $\omega$-regular expression.
\begin{alignat*}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
\end{alignat*}

\paragraph*{Result of simplification}
Those simplification rules lead to more intuitive representation of timelines. Here we demonstrate their effects using an example.

@@

    Using our algorithm, the \ltl\ formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates un-simplified $\omega$-regular expression $((\lnot a) | (aa^*(\lnot a)))((\lnot a) | (aa^*(\lnot a)))^{\omega}$, and simplified version $((\lnot a) | (aa^*(\lnot a)))^{\omega}$, which corresponds to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.

@@

\begin{remark}
    Both simplified and un-simplified $\omega$-regular expression would generate correct timeline representations. Correct as they faithfully represent the set of satisfying traces of the original \ltl\ formula $\Phi$. Nonetheless, we, as human users, oftentimes find the simplified version more intuitive to reason about.
\end{remark}

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz~\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, $A_{2i-1}$ could be $\epsilon$, and $\epsilon \not\in L(A_{2i})$. On a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. \cref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a1) {$A_1$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a2g) [right= 0.85cm of a1] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a2) [right= 1cm of a1] {$A_2$};
    \node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a3) [below= 0.4cm of a1] {$A_3$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a4g) [right= 0.85cm of a3] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a4) [right= 1cm of a3] {$A_4$};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a5) [below= 0.4cm of a3] {...};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a7) [below= 0.4cm of a5] {\!\!\!$A_{2n-1}$\!\!\!};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a8g) [right= 0.85cm of a7] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a8) [right= 1cm of a7] {\!\!\!$A_{2n}$\!\!\!};
    \node (m) at ($(a1)!0.5!(a7)$) {};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (s) [left= 1.5cm of m] {\!\!\!\!start\!\!\!\!};
    \draw[thick] (s) edge[-latex] (a1) (s) edge[-latex] (a3) (s) edge[-latex] (a5) (s) edge[-latex] (a7);
    \draw[thick] (a1) edge[-latex] (a2) (a3) edge[-latex] (a4) (a7) edge[-latex] (a8);
    \end{tikzpicture}
%    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}

@@

\subsection{Correctness of Timeline Visualization}\label{sec:correctness-timeline}
Inductively, we show how to construct each form of regular and $\omega$-regular expression into a timeline representation.
\paragraph{Regular expression}
\begin{enumerate}
  \item $A = \emptyset$. Visualized as no timeline. An empty regular expression signifies nothing is true at any time.
  \item $A = \epsilon$. Omitted in the visualization.
  \item $A = a$. Visualized as a single timeline node with $a$. A regular expression of an propositional logic formula signifies that the atomic proposition is true at the current time step.
  \item $A = A_1A_2$. Visualized as the timeline of $A_1$ followed by the timeline of $A_2$. A concatenation of two regular expressions signifies that the first regular expression must be true before the second regular expression can be true in the time step.
  \item $A = A_1 + A_2$. Visualized as the timelines of $A_1$ and $A_2$ in parallel. A union of two regular expressions signifies that either of the regular expressions can be true, which is represented by the parallel timelines.
  \item $A = A_1^*$. If $A_1$ is a propositional logic formula, then it is visualized as an egg shape node with caption ``repeats 0 - $\infty$''. If $A_1$ is a regular expression involving concatenation, star, etc, then it is visualized as the timeline of $A_1$ followed by a node with label ``$\cdots$'', and that timeline pattern again. A Kleene star of a regular expression signifies that the regular expression can be repeated any number of times, including 0 times, which matches the meaning of our representation of egg shape node and ``$\cdots$''  node as described in \cref{sec:timeline}.
\end{enumerate}
\paragraph{$\omega$-regular repression}
\begin{enumerate}
  \item $B = A^\omega$. Visualized as a grey box enclosed with the timeline of $A$. An $\omega$ of a regular expression signifies that the regular expression is to be repeated infinitely many times, which matches the meaning of our representation of grey box, because every time it reach the end of the grey box, it must reenter the same grey box again, hence infinite repeat.
  \item $B = AB$. Visualized as the timeline of $A$ followed by the timeline of $B$. A concatenation of a regular expression and an $\omega$-regular expression signifies that the first must be true before the second $\omega$-regular expression can be true in the time step.
  \item $B = B + B$. Visualized as the timelines of $B$ and $B$ in parallel. A union of two $\omega$-regular expressions signifies that either of the $\omega$-regular expressions can be true, which is represented by the parallel timelines.
\end{enumerate}

@@

    This formula is reasonably complicated, and hard for human to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.

>>>>

\subsection{Timeline}\label{sec:timeline}
We present timelines as graphic visualizations. It contains the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start''.
    \item Every node represents one time step, and each node has a propositional logic formula $\Psi$, which specifies the behavior of atomic propositions at that time step. The formula $\Psi$ must be true at that time step. If $\Psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item A node in egg shape with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting points.
    \item Timelines can be in parallel, signifying that either of the parallel timeline could happen.
\end{itemize}

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualizations (depicted in \Cref{fig:algo}) works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure*}

@@

\paragraph*{Rule-based simplification}
Here we show a demonstrating subset of the simplification rules we encoded. In theory one could add more rules to the tool, so long as they are sound; here we only choose to encode the rules that represent common patterns we have observed in the generated $\omega$-regular expressions.
\begin{alignat*}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
\end{alignat*}

\paragraph*{Result of simplification}
These simplification rules lead to more intuitive representation of timelines. Here we demonstrate their effects using an example.

@@

    Using our algorithm, the \ltl\ formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates the un-simplified $\omega$-regular expression $((\lnot a) | (aa^*(\lnot a)))((\lnot a) | (aa^*(\lnot a)))^{\omega}$, and the simplified version $((\lnot a) | (aa^*(\lnot a)))^{\omega}$, which correspond to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.

@@

\begin{remark}
    Both simplified and un-simplified $\omega$-regular expressions would generate correct timeline representations. Correct as they faithfully represent the set of satisfying traces of the original \ltl\ formula $\Phi$. Nonetheless, we, as human users, oftentimes find the simplified version more intuitive to reason about.
\end{remark}

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz~\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, $A_{2i-1}$ could be $\epsilon$, and $\epsilon \not\in L(A_{2i})$. At a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. \cref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a1) {$A_1$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a2g) [right= 0.85cm of a1] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a2) [right= 1cm of a1] {$A_2$};
    \node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a3) [below= 0.4cm of a1] {$A_3$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a4g) [right= 0.85cm of a3] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a4) [right= 1cm of a3] {$A_4$};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a5) [below= 0.4cm of a3] {...};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a7) [below= 0.4cm of a5] {\!\!\!$A_{2n-1}$\!\!\!};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a8g) [right= 0.85cm of a7] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a8) [right= 1cm of a7] {\!\!\!$A_{2n}$\!\!\!};
    \node (m) at ($(a1)!0.5!(a7)$) {};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (s) [left= 1.5cm of m] {\!\!\!\!start\!\!\!\!};
    \draw[thick] (s) edge[-latex] (a1) (s) edge[-latex] (a3) (s) edge[-latex] (a5) (s) edge[-latex] (a7);
    \draw[thick] (a1) edge[-latex] (a2) (a3) edge[-latex] (a4) (a7) edge[-latex] (a8);
    \end{tikzpicture}
%    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}

@@

\subsection{Correctness of Timeline Visualization}\label{sec:correctness-timeline}
Inductively, we show how to construct each form of regular and $\omega$-regular expression into a timeline representation.
\paragraph{Regular expression}
\begin{enumerate}
  \item $A = \emptyset$. Visualized as no timeline. An empty regular expression signifies nothing is true at any time.
  \item $A = \epsilon$. Omitted in the visualization.
  \item $A = a$. Visualized as a single timeline node with $a$. A regular expression of an propositional logic formula signifies that the atomic proposition is true at the current time step.
  \item $A = A_1A_2$. Visualized as the timeline of $A_1$ followed by the timeline of $A_2$. A concatenation of two regular expressions signifies that the first regular expression must be true before the second regular expression can be true in the time step.
  \item $A = A_1 + A_2$. Visualized as the timelines of $A_1$ and $A_2$ in parallel. A union of two regular expressions signifies that either of the regular expressions can be true, which is represented by the parallel timelines.
  \item $A = A_1^*$. If $A_1$ is a propositional logic formula, then it is visualized as an egg shape node with caption ``repeats 0 - $\infty$''. If $A_1$ is a regular expression involving concatenation, star, etc, then it is visualized as the timeline of $A_1$ followed by a node with label ``$\cdots$'', and that timeline pattern again. A Kleene star of a regular expression signifies that the regular expression can be repeated any number of times, including 0 times, which matches the meaning of our representation of egg shape node and ``$\cdots$''  node as described in \cref{sec:timeline}.
\end{enumerate}
\paragraph{$\omega$-regular repression}
\begin{enumerate}
  \item $B = A^\omega$. Visualized as a grey box enclosed with the timeline of $A$. An $\omega$ of a regular expression signifies that the regular expression is to be repeated infinitely many times, which matches the meaning of our representation of grey box, because every time it reach the end of the grey box, it must reenter the same grey box again, hence infinite repeat.
  \item $B = AB$. Visualized as the timeline of $A$ followed by the timeline of $B$. A concatenation of a regular expression and an $\omega$-regular expression signifies that the first must be true before the second $\omega$-regular expression can be true in the following timesteps.
  \item $B = B + B$. Visualized as the timelines of $B$ and $B$ in parallel. A union of two $\omega$-regular expressions signifies that either of the $\omega$-regular expressions can be true, which is represented by the parallel timelines.
\end{enumerate}

@@

    This formula is reasonably complicated, and hard for humans to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.
